The Stakes of Scrying: \texttt{\string.\string^} dotket and the Seer Proposal

### 2024-01-16 mss.tex
Revise.

\begin{abstract}
Nock 12 and the \dotket~rune break the referential transparency of Nock, with consequences for the purity of Urbit's functional programming model.  We consider the implications of this, and propose a move-based replacement for \dotket~that would restore referential transparency to Nock.
\end{abstract}

@@

The Nock 12 opcode and the \dotket~rune are used by userspace code to ask the kernel for data in the scry namespace.\footnote{The scry namespace, in brief, is a TODO bound}  When Urbit runs userspace code, that code is not raw Nock.  It is instead called Mock, which is Nock with two additions:   deterministic errors are caught and returned as values to the kernel; and there is an extra Nock opcode, opcode 12.  A simple Nock 12 formula looks like \texttt{\string[12 \string[1 /foo/bar/baz\string]\string]}, where \texttt{/foo/bar/baz} is a path literal \texttt{\string[\%foo \%bar \%baz \textasciitilde\string]}, representing a path in the scry namespace.  The result of running this opcode will be the result of performing the scry request, i.e. dereferencing the path.

@@

Userspace code cannot memoize Nock persistently because it is not feasible for the runtime to prevent a jet mismatch in the metacircular evaluator gate.  A gate like \texttt{++mink} takes in a Mock expression (a cell of \texttt{\string[subject formula\string]}) and a "scry handler gate" (which is described shortly), and returns a value which is one of:

@@

\begin{itemize}
  \item  \texttt{\string[\textasciitilde \textasciitilde value\string]} positive scry response;
  \item  \texttt{\string[\textasciitilde \textasciitilde\string] }permanent scry failure; or
  \item  \texttt{\textasciitilde} ``block''.
\end{itemize}

@@

The Nock 12 opcode is problematic because it renders the result of the computation dependent on external data which are not in the \texttt{\string[subject formula\string]} that is being evaluated.  The result is no longer a pure function of the \texttt{subject} and \texttt{formula}:  depending on what the scry handler gate is, the same \texttt{\string[subject formula\string]} could yield different results.  Consider \texttt{\string[subject formula\string]} of \texttt{\string[0 \string[12 1 /foo/bar\string]\string]}.  This returns the result of looking up the \texttt{/foo/bar} scry request path from the scry gate. Userspace code (which is untrusted) could call \texttt{(mink \string[\string[0 12 1 /foo/bar\string]\string] |=(\string^ \string[\textasciitilde \textasciitilde 33\string]))}, acquire the result \texttt{33}, then call \texttt{(mink \string[\string[0 12 1 /foo/bar\string]\string] |=(\string^ \string[\textasciitilde \textasciitilde 44\string]))} to yield \texttt{44}.  The \texttt{subject}s and \texttt{formula}s were identical, but because they were evaluated with different scry gates, they yielded different results.

@@

You could alternatively think of the scry namespace as an extra implicit argument passed to every Mock computation along with the subject and formula.  This extra argument is discovered dynamically over time as more ships bind more values, but since any path can only be bound once, it's immutable and all Mock expressions that dereference some scry path will either fail to complete or complete with identical results, hence pure functional.

>>>>

\begin{abstract}
Nock 12 and the \dotket~rune break the referential transparency of Nock, with consequences for the purity of Urbit's functional programming model.  We consider the implications of this, and propose a move-based replacement for the \dotket~rune that would restore referential transparency to Nock.
\end{abstract}

@@

The Nock 12 opcode and the \dotket~rune are used by userspace code to ask the kernel for data in the scry namespace.\footnote{The scry namespace, in brief, is a collection of immutable values “bound” by convention at fixed endpoints.  The scry namespace is defined by a unique path for each resource.  A representative scry path would have the entries \texttt{/host/rift/life/vane/request-type/revision-number/desk/file-path/mark}.}  When Urbit runs userspace code, that code is not raw Nock.  It is instead called Mock, which is Nock with two additions:   deterministic errors are caught and returned as values to the kernel; and there is an extra Nock opcode, opcode 12.  A simple Nock 12 formula looks like \texttt{\string[12 \string[1 /foo/bar/baz\string]\string]}, where \texttt{/foo/bar/baz} is a path literal \texttt{\string[\%foo \%bar \%baz \textasciitilde\string]}, representing a path in the scry namespace.  The result of running this opcode will be the result of performing the scry request, i.e. dereferencing the path.

@@

Userspace code cannot memoize Nock persistently because it is not feasible for the runtime to prevent a jet mismatch in the metacircular evaluator gate.  A gate like \texttt{++mink} takes in a Mock expression (a cell of \texttt{\string[subject formula\string]}) and a “scry handler gate” (which is described shortly), and returns a value which is one of:

@@

\begin{itemize}
  \item  \texttt{\string[\textasciitilde~\textasciitilde~value\string]} positive scry response;
  \item  \texttt{\string[\textasciitilde~\textasciitilde\string] }permanent scry failure; or
  \item  \texttt{\textasciitilde} ``block''.
\end{itemize}

@@

The Nock 12 opcode is problematic because it renders the result of the computation dependent on external data which are not in the \texttt{\string[subject formula\string]} that is being evaluated.  The result is no longer a pure function of the \texttt{subject} and \texttt{formula}:  depending on what the scry handler gate is, the same \texttt{\string[subject formula\string]} could yield different results.  Consider \texttt{\string[subject formula\string]} of \texttt{\string[0 \string[12 1 /foo/bar\string]\string]}.  This returns the result of looking up the \texttt{/foo/bar} scry request path from the scry gate. Userspace code (which is untrusted) could call \texttt{(mink \string[\string[0 12 1 /foo/bar\string]\string] |=(\string^ \string[\textasciitilde~\textasciitilde~33\string]))}, acquire the result \texttt{33}, then call \texttt{(mink \string[\string[0 12 1 /foo/bar\string]\string] |=(\string^ \string[\textasciitilde~\textasciitilde~44\string]))} to yield \texttt{44}.  The \texttt{subject}s and \texttt{formula}s were identical, but because they were evaluated with different scry gates, they yielded different results.

@@

Another way to think of the scry namespace is as comprising an extra implicit argument passed to every Mock computation along with the subject and formula.  This extra argument is discovered dynamically over time as more ships bind more values.  Since any path can only be bound once, it becomes immutable; all Mock expressions that dereference some scry path will either fail to complete or complete with identical results, hence pure functional.
